---
title: "SeatSeekr"
author: "Jan Murer, jan.murer@stud.unilu.ch, 17-100-777"
output: html_document
date: "2024-05-08"
---

# Introduction

As a capstone project for the seminar "Data Mining for the Social Sciences in R" i created a small shiny application that takes the current location of an user and shows the nearest available public toilet. 

The application uses the OpenCage API to get coordinates from an address and data from the opendata.swiss API to get a list of all public toilettes in Lucerne.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load necessary libraries to set up wrapper functions to interact with API

```{r}
library(tidyverse)
library(here)
library(httr)
library(jsonlite)
```

# Set variables for API keys and API-base URLs

```{r}

# Load API key from .csv using here package. Define base URL for OpenCage API.

api_key_opencage <- read.csv(here("00_api_credentials", "api_credentials.csv")) %>%
  pull(geocoding_api_key)

base_url_opencage <- "https://api.opencagedata.com/geocode/v1/json"

```

# Define wrapper function to interact with OpenCage API

```{r}

# Wrapperfunction to get user coordinates using OpenCage API

get_user_coordinates <- function(user_location) {
  
  # Make a GET request
  
  geocoding_response <- httr::GET(url = base_url_opencage, query = list(q = user_location, key = api_key_opencage))
  
  # Check status code
  
  if (geocoding_response$status_code == 200) {
    
    # Convert response to json
    
    geocoding_response_json <- httr::content(geocoding_response, as = "text") %>%
      fromJSON()
    
    # Extract coordinates (latitude/longitude)
    
    user_lat <- geocoding_response_json$results$geometry$lat[1]
    user_lng <- geocoding_response_json$results$geometry$lng[1]
    
    # Create a dataframe named user_coordinates
    
    user_coordinates <- data.frame(latitude = user_lat, longitude = user_lng)
    
    return(user_coordinates)
    
  } else {
    
    stop("Request failed with status code ", geocoding_response$status_code)
 
     }
}

```
